---
title: "Gene Expression Differential Analysis based on Limma"
author: "Stefano Monti"
output:
  html_document:
    theme: united
    toc: yes
    code_folding: show
    css: "../style/BS831.css" 
---

```{r include=FALSE}
knitr::opts_chunk$set(message=FALSE, warning=FALSE, eval=TRUE)
devtools::load_all(".")
require(Biobase)
require(limma)
```

Here, we present a couple of simple examples of differential analysis
based on `limma`. In particular, we show how the design matrix can be
constructed using different 'codings' of the regression variables.
We also define a simple wrapper function that can help us remember the
different `limma` steps.

This document also provides examples of how to embed equations into an
R markdown document, and how to cross-reference those equations.

```{r eval=FALSE}
require(BS831)
require(Biobase)
require(limma)
```

### Load the Data

We start by loading and pre-processing the necessary expression dataset.

```{r load.data, echo=TRUE}
## let us load the ExpressionSet, an already simplified version of the
## ..breast cancer dataset
data(renamedBreastDB)

dat <- renamedBreastDB
head(pData(dat))
table(dat$diseaseState)
```

### Differential Analysis based on `Limma`
When the regression variable is categorical (binary in this case), we
can choose different (yet equivalent) 'codings'. In particular, we can
fit a standard model

$$
\begin{equation}
y = \beta_0 + \beta_1 X_{group},
\label{eq:lm1}
\tag{1}
\end{equation}
$$

where $X_{group} = 0,1,$ if the observation is from a nonbasal- or a
basal-type tumor, respectively. Alternatively, we can fit the following model

$$
\begin{equation}
y = \beta_{\not{basal}} X_{\not{basal}} + \beta_{basal}X_{basal},
\label{eq:lm2}
\tag{2}
\end{equation}
$$

where $X_{\not{basal}} = 1,0$, if the observation is from a nonbasal-
or a basal-type tumor, respectively. Conversely, $X_{basal} = 0,1$, if
the observation is from a nonbasal- or a basal-type tumor,
respectively.

The corresponding models in `limma` are defined based on the two
design matrices `design1`, for model $\eqref{eq:lm1}$, and `design2`,
for model $\eqref{eq:lm2}$, as follows.

```{r limma.design}
design1 <- model.matrix( ~ diseaseState, data = pData(dat) )
colnames(design1) <- c("baseline","basal")      # let's simplify column names  
design2 <- model.matrix( ~ 0 + dat$diseaseState )
colnames(design2) <- levels( dat$diseaseState ) # ditto

print(unique(design1)) # showing only one instance of each class
print(unique(design2)) # ditto
```

The different designs require slightly different approaches to model
fitting. In model $\eqref{eq:lm1}$, we want to extract the p-value
corresponding to the $\beta_{group}$ parameter. In model
$\eqref{eq:lm2}$, we care about the _difference_ $\beta_{basal} -
\beta_{\not{basal}}$.

```{r limma.fit}
## model (1)
fit1 <- lmFit(dat,design1) # fitting of linear model
head(fit1$coefficients)
fit1 <- eBayes(fit1)       # pooling of variance across like-genes
fit1table <- topTable(fit1,coef="basal",adjust="BH",number=Inf,sort.by="P")
head(fit1table)

## model (2)
fit2 <- lmFit(dat,design2) # fitting of linear model
head(fit2$coefficients)
## let's define the contrast matrix we need
contrast.matrix <- makeContrasts((basal-nonbasal),levels=design2)
fit2 <- contrasts.fit(fit2,contrast.matrix)
## notice that the contrast coefficient takes the same values as the basal coefficient in design1
head(fit2$coefficients)
fit2 <- eBayes(fit2)       # pooling of variance across like-genes
fit2table <- topTable(fit2,coef=1,adjust="BH",number=Inf,sort.by="P")
head(fit2table)
```

Regardless of the 'coding', the results are the same.

```{r limma.compare}
## compare the results from the two 'codings'
all.equal(fit2table[,"t"],fit1table[rownames(fit2table),"t"])
plot(fit2table[,"t"],fit1table[rownames(fit2table),"t"],xlab="design2",ylab="design1")
```

### Defining a wrapper function

Since running `limma` requires memorizing several commands that are
not easy to remember, here we define a simple wrapper function that
will relief us from remembering them all. Incidentally, this is also
(a simplified version of) the wrapper function defined in `CBMRtools`,
and it corresponds to the `design2`, model $\eqref{eq:lm2}$, shown above.

It should be noted that this simple wrapper function does not allow
you to specify additional covariates (confounders). In that case, you
will have to either define a more advanced wrapper (any volunteer?),
or revert back to the direct use of limma commands.

```{r limma.wrapper}
run_limma <- function(
    eset,                # the expression set
    treatment,           # pData variable name of the phenotype
    cond,                # label of the 'treatment' group
    control,             # label of the 'control' group
    sort.by="P",         # sort output by given criterion
    decreasing = FALSE,  # sort order
    cutoff = NA          # extract only genes above/below 'sort.by' cutoff
)
{
    if ( ncol(fData(eset))<1 ) 
        stop( "empty fData" )
    
    cat( "running limma\n" )
    test_name <- paste(control, "_vs_", cond, sep = "")
    treatmentvec <- pData(eset)[, treatment]
    eset <- eset[, treatmentvec %in% c(cond, control)]
    treatmentvec <- factor( pData(eset)[, treatment] )
    design <- model.matrix( ~ 0 + treatmentvec )
    colnames(design) <- levels( treatmentvec )
    fit <- lmFit(eset, design)
    command_str <- paste("makeContrasts(", "(", cond , "-", control, ")", 
                         ",levels = design)", sep = "")
    contrast.matrix <- eval( parse(text=command_str) ) 
    fit2 <- contrasts.fit( fit, contrast.matrix )
    fit2 <- eBayes(fit2)
    
    ## extract full table
    fit2.table <- topTable(fit2, coef=1, adjust="BH", number=length(fit2), sort.by=sort.by)

    ## extract genes above/below cutoff
    if( !is.na(cutoff) ) {
        if (decreasing) {
            fit2.table <- fit2.table[fit2.table[,sort.by] > cutoff,]
        }
        else {
            fit2.table <- fit2.table[fit2.table[,sort.by] < cutoff,]
        }
    }
    return(fit2.table)
}
```

Now, let us test the newly defined function on our data.

```{r run.wrapper}
fit3table <- run_limma(dat,treatment="diseaseState",cond="basal",control="nonbasal")
head(fit3table)
all.equal(fit2table[,"t"],fit3table[rownames(fit2table),"t"])
```

