---
title: "Dimensionality Reduction"
author: "Stefano Monti, Anthony Federico"
output:
  html_document:
    theme: united
    toc: yes
    code_folding: show
    css: "../style/BS831.css" 
---

We illustrate the use of dimensionality reduction methods to visualize
high-dimensional data into low-dimensional (2D or 3D) space, so as to
gain some intuition about their structure. A typical purpose of these
visualizations is to carry out quick QC, to evaluate wether the data
behaves as expected, or to identify (possibly unexpected) outliers or
anomalies.

```{r include=FALSE}
knitr::opts_chunk$set(message=FALSE, warning=FALSE, eval=TRUE)
devtools::load_all(".")
require(Biobase)
```

```{r eval=FALSE}
require(BS831)
require(Biobase)
```

```{r load.data}
data(DMsubset)

subset <- c("TESTOSTERONE","DOXORUBICIN","DIAZEPAM") 
dm3 <- dm10[,phenoData(dm10)$CHEMICAL %in% subset]             # subset data on the 3 chemicals
phenoData(dm3)$CHEMICAL <- droplevels(phenoData(dm3)$CHEMICAL) # eliminate unused levels from factor
col_palette <- c("red","dark green","blue")
dm3_col <- col_palette[phenoData(dm3)$CHEMICAL]
```

In this example, we show three commonly used dimensionality reduction
algorithm with our toy dataset.

## Principal Component Analysis (PCA)

### PCA with SVD

We start with Principle Component Analysis 'PCA', using the function
`prcomp`. This will perform PCA using Singular Value Decomposition. We
then print the summary statistics, which tell us the variance
explained by each component. We then visualize the dataprojected on
the first two components, and then projected on the 1st and 3rd
component.

```{r PCA, message=FALSE, warning=FALSE}
dm3_svd <- prcomp(Biobase::exprs(dm3), scale = TRUE)  ## perform PCA
summary(dm3_svd)$importance[,1:4]        ## show variance explained by each component

## let's plot the first two components
plot(x=dm3_svd$rotation[,"PC1"],y=dm3_svd$rotation[,"PC2"],col=dm3_col,
     main="PCA",xlab="1st PC", ylab="2nd PC",pch=20) 
legend("topleft",col=col_palette,legend=levels(phenoData(dm3)$CHEMICAL),pch=20)

## let's plot the first and third component
plot(x=dm3_svd$rotation[,"PC1"],y=dm3_svd$rotation[,"PC3"],col=dm3_col,
     main="PCA",xlab="1st PC", ylab="3rd PC",pch=20)
legend("topright",col=col_palette,legend=levels(phenoData(dm3)$CHEMICAL),pch=20)
```

Let us use the "fancier" plotting functions in the `ggplot` package.

```{r pca.ggplot, message=FALSE, warning=FALSE}
require(ggplot2)
df <- data.frame(x=dm3_svd$rotation[,"PC1"], y=dm3_svd$rotation[,"PC2"],
                 id=phenoData(dm3)$CHEMICAL)

p2 <- ggplot(df, aes(x=x, y=y, colour=id)) +
    geom_point() +
        theme_bw() +
            xlab("PC1") + ylab("PC2") + ggtitle("PCA 1 vs. PCA2")
p2
```

It appears that the first two components do a better job at separating
the three chemicals than the 1st and 3rd components. We can bring it
altogether in an interactive 3D plot.

```{r pca.3d, message=FALSE, warning=FALSE}
library(plotly)
library(magrittr)

df.pca <- cbind(dm3_svd$rotation[,c(1:3)], as.character(phenoData(dm3)$CHEMICAL)) %>%
          as.data.frame() %>%
          magrittr::set_colnames(c("PC1", "PC2", "PC3", "id"))

plot_ly(df.pca,
        x=~PC1,
        y=~PC2,
        z=~PC3,
        type="scatter3d",
        mode="markers",
        color=~id,
        marker=list(size=5))
```

### PCA with Eigen Decomposition

Next, we perform Principle Component Analysis 'PCA', using the function
`princomp`. This will perform PCA using eigen decomposition. We then visualize the data
projected on the first two components, and then projected on the 1st and 3rd component.

```{r PCA_eg, message=FALSE, warning=FALSE}
dm3_eg <- princomp(Biobase::exprs(dm3))  ## perform PCA
colnames(dm3_eg$loadings) <- paste0("PC", 1:length(colnames(dm3_eg$loadings)))

```

Again with `ggplot2`.

```{r pca.ggplot_eg, message=FALSE, warning=FALSE}
df <- data.frame(PC1=dm3_eg$loadings[,"PC1"], PC2=dm3_eg$loadings[,"PC2"], PC3=dm3_eg$loadings[,"PC3"],
                 id=phenoData(dm3)$CHEMICAL)

p2 <- ggplot(df, aes(x=PC1, y=PC2, colour=id)) +
    geom_point() +
        theme_bw() +
            xlab("PC1") + ylab("PC2") + ggtitle("PCA 1 vs. PCA2")
p2

p3 <- ggplot(df, aes(x=PC1, y=PC3, colour=id)) +
    geom_point() +
        theme_bw() +
            xlab("PC1") + ylab("PC3") + ggtitle("PCA 1 vs. PCA3")
p3
```

Again, it appears that the first two components do a better job at separating
the three chemicals than the 1st and 3rd components. Debatable whether separation with is better with ``prcomp`` (SVD) than ``princomp`` (eigen decomposition).

### t-Distributed Stochastic Neighbor Embedding (T-SNE)

Lastly, we carry out projection and visualization based on the <a
href="http://www.jmlr.org/papers/v9/vandermaaten08a.html">t-SNE
method</a> (see this nice <a href="http://distill.pub/2016/misread-tsne/
">tutorial</a>).

```{r t-SNE, message=FALSE, warning=FALSE}
require(tsne)

# Create distance matrix (Euclidean)
distDM3 <- dist(t(Biobase::exprs(dm3)))

set.seed(123) ## t-SNE with perplexity=10
subset_tsne <- tsne(distDM3,k=2,initial_dims=300,max_iter=3000,perplexity=3)
plot(x=subset_tsne[,1],y=subset_tsne[,2],col=dm3_col,main="perplexity=3",pch=20)

set.seed(123) ## t-SNE with perplexity=10
subset_tsne <- tsne(distDM3,k=2,initial_dims=300,max_iter=3000,perplexity=10)
plot(x=subset_tsne[,1],y=subset_tsne[,2],col=dm3_col,main="perplexity=10",pch=20)

set.seed(123) ## t-SNE with perplexity=10
subset_tsne <- tsne(distDM3,k=2,initial_dims=300,max_iter=3000,perplexity=20)
plot(x=subset_tsne[,1],y=subset_tsne[,2],col=dm3_col,main="perplexity=20",pch=20)

## here we show lack of "convergence" if not enough iterations are run
set.seed(123) ## t-SNE with perplexity=40
subset_tsne <- tsne(distDM3,k=2,initial_dims=300,max_iter=2000,perplexity=40)
plot(x=subset_tsne[,1],y=subset_tsne[,2],col=dm3_col,main="perplexity=30, 2K iterations",pch=20)

## convergence reached after 5000 iterations
set.seed(123) ## t-SNE with perplexity=30
subset_tsne <- tsne(distDM3,k=2,initial_dims=300,max_iter=5000,perplexity=30)
plot(x=subset_tsne[,1],y=subset_tsne[,2],col=dm3_col,main="perplexity=30, 5K iterations",pch=20)

## here, the perplexity is too large to yield good separation (also 
## ..tried w/ 10K iterations, not shown)
set.seed(123) ## t-SNE with perplexity=40
subset_tsne <- tsne(distDM3,k=2,initial_dims=300,max_iter=3000,perplexity=40)
plot(x=subset_tsne[,1],y=subset_tsne[,2],col=dm3_col,main="perplexity=40",pch=20)
```

### Uniform Manifold Approximation and Projection (UMAP)
```{r}
library(umap)

print(umap.defaults)

umap.custom <- umap.defaults
umap.custom$n_neighbors <- 3
umap.custom$n_components <- 3
umap.custom$metric <- "euclidean"

umap.data <- umap(t(Biobase::exprs(dm3)), config=umap.custom, method="naive")
df.umap <- umap.data$layout 
colnames(df.umap) <- paste("C", seq_len(ncol(df.umap)), sep="")
df.umap <- as.data.frame(df.umap)
df.umap$id <- as.character(phenoData(dm3)$CHEMICAL)

head(df.umap)

plot_ly(df.umap,
        x=~C1,
        y=~C2,
        z=~C3,
        type="scatter3d",
        mode="markers",
        color=~id,
        marker=list(size=5))
```

### Summary

- Use SVD (`prcomp`) to perform PCA over (`princomp`).
- tSNE is a useful way to visualize data.
    - Drawbacks: Need to set a seed to get reprocuible results.
    - Need to play with perplexity parameter in order to get good separation.
