---
title: "Probability Distributions"
author: "Stefano Monti"
output:
  html_document:
    theme: united
    toc: yes
    code_folding: show
    css: "../style/BS831.css"
---

Let us generate data and plot from few examples of probability
distributions. In general, any given distribution `dist` in R is
defined by four functions `rdist`, `pdist`, `ddist`, and `qdist`,
which correspond to the sampling, the CDF, the density (or mass) and
the quantile functions, respectively. For example, the Normal
distribution is defined by the four functions `rnorm`, `pnorm`,
`dnorm`, and `qnorm`.

Let us generate a sample (X) of 100 points from a standard normal
(mean 0 and stdev 1); visualize it as a boxplot and a histogram; generate a second
sample (Y) of points that are linearly related to the first sample (based on the 
linear equation $y = 2x + \epsilon$, with $\epsilon \sim N(0,1)$.

```{r norm0}
N <- 100 
X <- rnorm(N,mean=0,sd=1)
par(mfrow=c(1,2))
boxplot(X)
hist(X)
par(mfrow=c(1,1))

Y <- 2 + X + rnorm(N)
plot(X,Y,pch=20)
abline(2,1,col="red") # plot a line with intercept 2 and slope 1
abline(h=0,lty=3)     # show the origin
abline(v=0,lty=3)     # ..
```

Let us now plot the well-known bell-shaped curve and the corresponding CDF.

```{r plot1}
X <- seq(-4,4,.1)
plot(X,dnorm(X),pch=20)
plot(X,pnorm(X),pch=20)
```

### A simple distribution plot function

Notice that in the above code, the points plotted are evenly spaced,
which is not optimal, since we would like to have higher resolution
where most of the density lies. Furthermore, we need to a-priori know where
most of a distribution mass lies, so as to generate an adequate set of X coordinates. 
Otherwise, we might end up with a truncated plot, as shown in the examples below.

```{r plot2}
X <- seq(-4,4,.1)
plot(X,dnorm(X,mean=0,sd=3),pch=20) # too narrow range
plot(X,dnorm(X,mean=3,sd=1),pch=20) # range skewed to the left
```

To take care of this, we can define a slightly more sophisticated
distribution plotting function, using information about a
distribution's quantiles.

```{r init}
plot.distn <- function(qfun, dfun, n, type="l", lty=1, add=FALSE, xlab=NULL, ylab=NULL, main=NULL, col="black", silent=TRUE, ...)
{
  p <- (1:(n-1))/n
  q <- qfun(p,...) # generate the set of X-coordinates based on the distribution quantiles
  
  if ( add )
    lines( q, dfun(q,...), lty=lty, col=col )
  else
    plot( q, dfun(q,...), type=type, lty=lty, col=col, xlab=xlab,ylab=ylab,main=main )

  ## this is useful to determine plot boundaries
  if (!silent) return ( c(0,max(dfun(q,...))) )
}
```

### Normal (or Gaussian) distribution

Now we can plot a histogram of 1000 data points drawn from a standard
(i.e., mean=0, stdev=1) Normal (or Gaussian) distribution.

```{r norm1}
X1 <- rnorm(1000,mean=0,sd=1) 
hist(X1)
```

If we want the histogram to show frequencies (i.e., numbers in the 0-1
range), rather than counts, so that we can superimpose the density
contour (using the function `plot.distn`), we can use the argument
`probability=TRUE`

```{r norm2}
hist(X1,probability=TRUE)
plot.distn(qnorm,dnorm,n=1000,add=TRUE)

# let's increase sample size to increase histogram resolution
X1 <- rnorm(1000000,mean=0,sd=1) 
hist(X1,probability=TRUE)
plot.distn(qnorm,dnorm,n=1000,add=TRUE)
```

### Log-Normal distribution
Let's plot from a log-normal distribution.

```{r lnorm1}
N <- 1000 # number of datapoints
X2 <- rlnorm(N,meanlog=0,sdlog=1)
hist(X2,probability=TRUE)
plot.distn(qlnorm,dlnorm,n=N,add=TRUE,meanlog=0,sdlog=1)

# the plot is truncated. Here's a fix (won't always work)
plot.distn(qlnorm,dlnorm,n=N,meanlog=0,sdlog=1)
hist(X2,probability=TRUE,add=TRUE)

# a log-normal distribution yields data whose logarithm is normally distributed
X1 <- rnorm(N,mean=0,sd=1)       # remember X1 ~ N(0,1)
qqplot(X1,log(X2),main='qqplot') # plot it against X2 ~ log-N(0,1)
abline(0,1,col="red")

## qqplot is equivalent to ..
plot(sort(X1),sort(log(X2)))

## let's see how different means and standard deviations look like ...
X2 <- rlnorm(N,meanlog=0,sdlog=0.5)
plot.distn(qlnorm,dlnorm,n=10*N,meanlog=0,sdlog=0.5,ylab="dlnorm(q,...)")
hist(X2,probability=TRUE,add=TRUE)

plot.distn(qlnorm,dlnorm,n=10*N,meanlog=0,sdlog=0.5,ylab="dlnorm(q,...)")
plot.distn(qlnorm,dlnorm,n=10*N,meanlog=0,sdlog=1.0,ylab="dlnorm(q,...)",add=TRUE, col='red',lty=2)
plot.distn(qlnorm,dlnorm,n=10*N,meanlog=0,sdlog=5.0,ylab="dlnorm(q,...)",add=TRUE, col='blue',lty=3)
legend('topright',col=c('black','red','blue'),lty=1:3,legend=c("sdlog=0.5","sdlog=1.0","sdlog=5.0"))
```

