---
title: "Gene Set Projection based on KS"
author: "Stefano Monti"
output:
  html_document:
    theme: united
    toc: yes
    code_folding: show
    css: "../style/BS831.css" 
---

```{r include=FALSE}
knitr::opts_chunk$set(message=FALSE, warning=FALSE, eval=TRUE)
#devtools::load_all(".")
library(BS831)
library(GSVA)
library(Biobase)
library(cba)
library(ComplexHeatmap)
library(circlize)
```

In this module, we show how to use Kolmogorov-Smirnov (KS)-based
scoring to map gene expression data to geneset expression data. The
approach is useful to _abstract_ from the gene level to the signature
or pathway level, thus making the data more interpretable while at the
same time making them more robust to potential batch-to-batch effects.

```{r eval=FALSE}
library(BS831)
library(GSVA)
library(Biobase)
library(cba)
library(ComplexHeatmap)
library(circlize)
```

We start by requiring the necessary package and loading the data. The
data consists of two objects: a "toy" expression dataset of 126 genes
(`gsp.eSet`); and a list of four genesets (`gsp.GeneSet`).

```{r load_data}
data(gspData)

## make sure the expected objects have been uploaded ..
if (is.null(gsp.eSet)) stop("is.null(gsp.eSet)")
if (is.null(gsp.GeneSet)) stop("is.null(gsp.GeneSet)")

## show the size of the genesets and of the expression dataset
print(sapply(BS831::getGeneSet(gsp.GeneSet),length))
print(dim(gsp.eSet))
```

We then rename the dataset rows, so as to have them indexed by gene symbols.

```{r rename}
DAT1 <- gsp.eSet[fData(gsp.eSet)[,"symbol"]!="",]     # eliminate rows w/o gene symbols
Biobase::exprs(DAT1) <- log2(Biobase::exprs(DAT1)+1)  # log-transform the data
featureNames(DAT1) <- toupper(fData(DAT1)[,"symbol"]) # replace row names
DAT1$my_grade <- factor(DAT1$my_grade)
DAT1$my_stage <- factor(DAT1$my_stage)
```

We next visualize the raw, gene-level, data. We use the heatmap
function avaliable from *heatmap.ggplot.R*.

```{r visualize_raw, message=FALSE, warning=FALSE, results='hide'}
grade_col <- rainbow(n=nlevels(DAT1$my_grade)); names(grade_col) <- levels(DAT1$my_grade)
stage_col <- rainbow(n=nlevels(DAT1$my_stage)); names(stage_col) <- levels(DAT1$my_stage)
ha.genes <- HeatmapAnnotation(grade=DAT1$my_grade,
                              stage=DAT1$my_stage,
                              col=list(grade=grade_col,stage=stage_col))

Heatmap(t(scale(t(exprs(DAT1)))),
        name="gene expression", 
        col=colorRamp2(c(-3, 0, 3), c("#072448", "white", "#ff6150")),
        top_annotation=ha.genes, 
        cluster_rows=TRUE,
        cluster_columns=TRUE,
        clustering_distance_rows="euclidean",
        clustering_method_rows="ward.D",    
        clustering_distance_columns="euclidean",
        clustering_method_columns="ward.D", 
        show_parent_dend_line=TRUE,
        row_title="",
        show_column_names=FALSE,
        show_row_names=FALSE)
```

We next use the package `gsva` to perform single sample geneset projection.

```{r runGSP, warning=FALSE}
GSPdir <- gsva(DAT1, getGeneSet(gsp.GeneSet), mx.diff=FALSE, verbose=FALSE, parallel.sz=1)
## extract the ExpressionSet
print(dim(GSPdir))
```

.. and visualize the results. In this toy example, we projected onto four genesets only. In reality, one can use an entire geneset compendium (e.g., all genesets in MSigDB's
<a href="http://software.broadinstitute.org/gsea/msigdb/genesets.jsp?collection=CP">c2.cp</a>
compendium) to obtain a richer output.

```{r visualize, message=FALSE, warning=FALSE,results='hide'}
featureNames(GSPdir) <- c("WNT_BCATENIN","BCATENIN.CSNK1E","BCATENIN.CTBP1","TGF_BETA")

Heatmap(t(scale(t(exprs(GSPdir)))),
        name="gene expression", 
        col=colorRamp2(c(-3, 0, 3), c("#072448", "white", "#ff6150")),
        top_annotation=ha.genes, 
        cluster_rows=TRUE,
        cluster_columns=TRUE,
        clustering_distance_rows="euclidean",
        clustering_method_rows="ward.D",    
        clustering_distance_columns="euclidean",
        clustering_method_columns="ward.D", 
        show_parent_dend_line=TRUE,
        row_title="",
        show_column_names=FALSE,
        show_row_names=TRUE)
```

