---
title: "Mapping genesets for KS testing"
author: "Amy Li, Eric Reed, Stefano Monti"
output:
  html_document:
    theme: united
    toc: yes
    code_folding: show
    css: "../style/BS831.css" 
---

```{r include=FALSE}
knitr::opts_chunk$set(message=FALSE, warning=FALSE, eval=TRUE)
devtools::load_all(".")
ngenes <- 10
```

```{r eval=FALSE}
require(BS831)
ngenes <- 10
```

Let us generate a toy set of `r ngenes` genes with random tScores,
which we will use to establish a gene ranking and based on
which we will perform the KS enrichment.

```{r t.scores}
set.seed(123)
tScores <- round(rnorm(ngenes,0,1),4)
names(tScores) <- sprintf("gene%02d",1:ngenes)
print(tScores)
```

Notice that the t.score vector is named (i.e., each score is
labeled by the corresponding gene).

Let's assume the geneset we want to test is the following:

```{r gset}
gset <- c("gene03","gene05","gene06","gene07")
```

To establish the geneset members' ranking, and to properly perform the
KS test, we can proceed in one of two ways.

1. We can sort the t.score vector, find the position of the geneset
members and call ksGenescore on those positions

```{r sort}
## sorting the negative of the scores yields decreasing order sort
## alternatively: sortT <- tScores[order(tScores,decreasing=TRUE)]
sortT <- sort(-tScores)

## find the geneset members' positions
idx1 <- match(gset,names(sortT))
print(idx1) # show those positions (toward the end of the ranking)

## perform ks test
ksGenescore(ngenes,idx1,do.plot=TRUE,main="by sort")
```

2. We can compute the rank of the tScores, find the ranks of the
geneset members, and call ksGenescore on those ranks

```{r rank}
## computing ranks of minus the scores (to use decreasing order)
rankT <- rank(-tScores,ties.method="first")

## find the positions of geneset members in the rank list
idx2 <- match(gset,names(rankT))
print(rankT[idx2]) # show those ranks

## perform ks test on the ranks
ksGenescore(ngenes,rankT[idx2],do.plot=TRUE,main="by rank")
```


