---
title: "Heatmaps"
author: "Stefano Monti, Anthony Federico, Amy Li"
output:
  html_document:
    theme: united
    toc: yes
    code_folding: show
    css: "../style/BS831.css" 
---

```{r include=FALSE}
knitr::opts_chunk$set(message=FALSE, warning=FALSE, eval=TRUE)
devtools::load_all(".")
library(pheatmap)
library(Biobase)
library(ggdendro)
library(scales)
library(cba)
library(gtable)
library(gridExtra)
library(ComplexHeatmap)
library(circlize)
```

```{r eval=FALSE}
library(BS831)
library(cba)
##library(heatmap.plus)
library(pheatmap)
library(Biobase)
library(ggdendro)
library(scales)
library(gtable)
library(gridExtra)
library(ComplexHeatmap)
library(circlize)
```

## Visualization of a Gene Expression Dataset: the Drug Matrix 

We will use the DrugMatrix subset previously described for our
examples. This data corresponds to chemical perturbation experiments,
whereby rats are exposed to different chemicals, and their liver's RNA
is profiled. The chemicals are annotated both in terms of their
carcinogenicity and their genotoxicity, and we will use these two
phenotypes to color-code the samples in the visualized heatmaps.

```{r drugMatrix, message=FALSE, warning=FALSE}
data(dm10)
DM <- variationFilter(dm10, ngenes=500,do.plot=FALSE)
```

## The Function `pheatmap`

We first illustrate the use of the function `pheatmap` from the omonimous package.  This function only requires a numeric matrix as input, and it allows for annotation of columns and rows.  It can perform different clustering methods on rows an columns, either by specifying parameters of the clustering method to use, or by inputting the output of free-standing clustering functions such as `hclust`. Check `?pheatmap` to explore different arguments you can set, there are a lot of them.

Let us first define a simple function to create a color gradient to be used for coloring the gene expression heatmaps.

```{r colGradient, message=FALSE, warning=FALSE}
colGradient <- function( cols, length, cmax=255 )
{
  ## e.g., to create a white-to-red gradient with 10 levels
  ##
  ##   colGradient(cols=c('white','red'),length=10)
  ##
  ## or, to create a blue-to-white-to-red gradients with 9 colors (4 blue's, white, 4 red's)
  ##
  ##   colGradient(cols=c('blue','white','red'),length=9)
  ##
  ramp <- grDevices::colorRamp(cols)
  rgb( ramp(seq(0,1,length=length)), max=cmax )
}
```

We can then establish the color coding for the expression levels (blue=low, white=medium, to red=high).

```{r heatmap.colgradient}
## color gradient for the expression levels (blue=down-regulated; white=neutral; red=up-regulated)
bwrPalette <- colGradient(c("blue","white","red"),length=11)
```

Then, we will format the phenotype annotation for the columns of the heatmap, as well as specify the colors for each category for each variable, i.e., carcinogenicity (NA/negative/positive) and genotoxicity (NA/0/1), for both of which we will use the same color-coding, and chemicals. Notice the use of the function `rainbow` to generate a large number of colors for the annotaiton of the chemicals.

```{r sample.color.coding}
## color coding of the samples indicating carcinogenicity and genotoxicity 
chemPalette <- c("white","green","purple") # we will use the same colors for carc and gtox

annot <- pData(DM)[, c("Carcinogen_liv", "GenTox", "CHEMICAL")]
colnames(annot) <- c("CARC", "GTOX", "CHEM")
annot$CHEM[annot$CHEM == ""] <- NA
annot$CHEM <-  factor(annot$CHEM, levels = levels(annot$CHEM)[levels(annot$CHEM)!=""])
annot$GTOX <- as.factor(annot$GTOX)

annotCol <- list(
  CARC = chemPalette[-1],
  GTOX = chemPalette[-1],
  CHEM = rainbow(length(levels(annot$CHEM)))
)
names(annotCol$CARC) <- c("NON-CARC","CARCINOGEN")
names(annotCol$GTOX) <- c("0","1")
names(annotCol$CHEM) <- levels(annot$CHEM)
```

Next, we will create the heatmap, given the expression matrix from DrugMatrix.  Note that the argument `scale = "row"`, specifies that we want to scale each row (or gene) to mean zero, and let the colors denote the number of standard deviations from the mean. Below, we show the hetamaps with a couple of different codings of the color gradient, one using `grDevices::colorRamp` utilized in the function `colGraident`, and the other using `circlize::colorRamp2`. This is just to show there are multiple ways you can specify a color gradient.

## {.tabset}

### `grDevices::colorRamp`

```{r pheatmap.colorRamp, message=FALSE, warning=FALSE, fig.align="center", fig.width=10}
pheatmap(exprs(DM),
        color=bwrPalette, 
        annotation_col = annot,
        annotation_colors = annotCol,
        clustering_method = "ward.D", # default is 'complete'
        show_rownames = FALSE,
        show_colnames = FALSE,
        scale = "row")
```

### `circlize::colorRamp2`

```{r pheatmap.colorRamp2, message=FALSE, warning=FALSE, fig.align="center", fig.width=10}
pheatmap(exprs(DM),
        color=circlize::colorRamp2(c(-3, 0, 3), c("#072448", "white", "#ff6150")), 
        annotation_col = annot,
        annotation_colors = annotCol,
        clustering_method = "ward.D", # default is 'complete'
        show_rownames = FALSE,
        show_colnames = FALSE,
        scale = "row")
```

## {-}

## The package `ComplexHeatmap`

Here, we illustrate the use of Bioconductor package [`ComplexHeatmap`](https://bioconductor.org/packages/release/bioc/html/ComplexHeatmap.html), one of the most recently developed and most versatile. Its functionalities go well beyond the ones illustrated here.

Here, we show the heatmap ordered by hierarchical clustering (both rows and columns), with the columns split into the four main clusters (as would be determined by `cutree(...,k=4)`). As explained in the [package's manual](https://jokergoo.github.io/ComplexHeatmap-reference/book/a-single-heatmap.html#split-by-categorical-variables), "_[i]f `row_km/column_km` is set or `row_split/column_split` is set as a vector or a data frame, hierarchical clustering is first applied to each slice which generates k dendrograms, then a parent dendrogram is generated based on the mean values of each slice._"

```{r ComplexHeamap}
## Scale expression by row 
print(BS831::scale_row)
DMscaled <- BS831::scale_row(DM)

# Take columns you want from phenotype data
ha.t <- HeatmapAnnotation(CHEMICAL=DMscaled$Carcinogen_liv,
                          GTOX=as.factor(DMscaled$GenTox),
                          na_col="ghostwhite",
                          col=list(CHEMICAL=c("NON-CARC"="green",CARCINOGEN="red"),
                                   GTOX=c('0'="orange",'1'="purple")))

Heatmap(Biobase::exprs(DMscaled),
        name="expression", 
        col=circlize::colorRamp2(c(-3, 0, 3), c("#072448", "white", "#ff6150")),
        top_annotation=ha.t, 
        cluster_rows=TRUE,
        cluster_columns=TRUE,
        clustering_distance_rows="euclidean",
        clustering_method_rows="ward.D",    
        clustering_distance_columns="euclidean",
        clustering_method_columns="ward.D", 
        column_split=4,
        show_parent_dend_line=TRUE,
        row_title="",
        show_column_names=FALSE,
        show_row_names=FALSE)
```

In the following example, we use k-means clustering instead to split the dataset into four clusters (`column_km=4`) and then we perform hierarchical clustering within each cluster. To ensure reproducibility of results (since k-means has a random component), we set the random seed first.

```{r ComplexHeatmap.kmeans}
set.seed(123) ## for reproducibility of the k-means results
Heatmap(Biobase::exprs(DMscaled),
        name="expression", 
        col=circlize::colorRamp2(c(-3, 0, 3), c("#072448", "white", "#ff6150")),
        top_annotation=ha.t, 
        cluster_rows=TRUE,
        cluster_columns=TRUE,
        clustering_distance_rows="euclidean",
        clustering_method_rows="ward.D",    
        clustering_distance_columns="euclidean",
        clustering_method_columns="ward.D", 
        column_km=4, 
        show_parent_dend_line=TRUE,
        row_title="",
        show_column_names=FALSE,
        show_row_names=FALSE)
```

