---
title: "Efficient Computation in R"
author: "Stefano Monti"
output:
  html_document:
    theme: united
    toc: yes
    code_folding: show
    css: "../style/BS831.css" 
---

```{r global, echo=FALSE}
knitr::opts_chunk$set(message=FALSE, warning=FALSE)
```

Here we show a simple example of the gains in efficiency that can be achieved by taking advantage of matrix operations rather than multiple calls to a function. For more on the topic, see [this RCS module](https://drive.google.com/file/d/1cObV9rrXIi_vyGtKer6pd6mRz_347_m6/view?usp=share_link).

In this example, we compare three approaches to computing the row sums
of a large matrix. In the first (inefficient) approach, we make
use of a double `for` loop. In the second (still inefficient) approach, we
make use of the function <code>apply</code>. In the third (most
efficient) approach, we make use of the cartesian product operator
`%*%`.

## Use of `for` loop
```{r rowSum}
set.seed(123) # for reproducible results
N <- 10000000 # 10^7
X <- matrix(rnorm(N),nrow=N/10,ncol=10) # a 10^6-by-10 matrix

## use of for loop
forSum <- function(X)
{
    rowS <- rep(0,nrow(X))
    for ( i in 1:nrow(X) )
        for ( j in 1:ncol(X) )
            rowS[i] <- rowS[i] + X[i,j]
    return( rowS )
}
Tfor <- system.time(tmp0 <- forSum(X))
print(Tfor)
```

## Use of `apply`
```{r rowSum.apply}
## use of apply function
Tsum <- system.time(tmp1 <- apply(X,1,sum))
print(Tsum)
```

## Use of inner product (`%*%`)
```{r rowSum.inner}
## use of matrix multiplication
In <- rep(1,10) # define a unit vector
Tprd <- system.time(tmp2 <- X %*% In)
print(Tprd)

## let's measure the speed-up
Tsum["sys.self"]/Tprd["sys.self"] # CPU
Tsum["user.self"]/Tprd["user.self"] # CPU + R/W
Tfor["user.self"]/Tprd["user.self"] # CPU + R/W
```

As you can see, the matrix-based sum achieves a `r round(Tsum["user.self"]/Tprd["user.self"])`-fold speed-up relative to the `apply`-based implementation, and a `r round(Tfor["user.self"]/Tprd["user.self"])`-fold speed-up relative to the `for loop`-based implementation.]

It should be noted that there are actually native R functions for the
efficient sum of a matrix rows and columns (`rowSums` and
`colSums`). However, the point of this basic demonstration holds.

The gain in efficiency is due to the fact that in the `for` loop formulation, all the operations are performed in R (which is an interpreted language, hence inefficient), and in the `apply` formulation there are 1M function calls (1M calls of the `sum` function), which is also inefficient. In the `%*%` formulation, there is a single R function call, and all the number crunching is performed in whatever (compiled) language the `%*%` is implemented, most likely Fortran.

In fact, notice that the gain in efficiency of the `%*%` over `apply` is largely lost when computing the column sum, since there are only 10 columns, hence only ten calls of the `sum` function. The gain in efficiency with respect to the `for` loop is still there.

```{r colSum.for}
## use of apply function
T3 <- system.time(tmp3 <- forSum(t(X)))
print(T3)
```

```{r colSum.apply}
## use of apply function
T4 <- system.time(tmp4 <- apply(X,2,sum))
print(T4)
```

```{r colSum.inner}
## use of matrix multiplication
In <- rep(1,1000000) # define a unit vector
T5 <- system.time(tmp5 <- t(X) %*% In)
print(T5)
```

## Native row- (and col-)wide R functions
As noted, cognizant of this issue, R provides efficient functions to perform row and column sums. 

```{r rowSums}
T6 <- system.time(rowSums(X))
print(T6)
```

```{r colSums}
T7 <- system.time(colSums(X))
print(T7)
```

