---
title: "Multiple-Hypothesis Correction"
author: "Stefano Monti"
output:
  html_document:
    theme: united
    toc: yes
    code_folding: show
    css: "../style/BS831.css" 
---

In this module, we show how testing for multiple hypotheses (genes) can
increase the chance of false positives, especially for small sample
sizes. 

```{r include=FALSE}
knitr::opts_chunk$set(message=FALSE, warning=FALSE, eval=TRUE)
devtools::load_all(".")
library(heatmap.plus)
library(ggplot2)
library(tidyr)
```

```{r eval=FALSE}
library(BS831)
library(heatmap.plus)
library(ggplot2)
library(tidyr)
```

# Multiple-Hypothesis Correction by FDR and FWER

In the remainder of this module we will explore the effects of
multiple hypothesis corrections on family-wise error rate, false
positive rate, false negative rate, and power when using <a
href="https://en.wikipedia.org/wiki/Bonferroni_correction">Bonferroni</a>
and <a href="https://www.jstor.org/stable/2346101">Benjamini
Hochberg</a> corrections.

## Define functions

The following function is just a wrapper for runnning two-sample t-tests.

### Wrapper of Student's T-Test

```{r diffanal}
## wrapper function extracting t-statistic and p-value from a call to function t.test
tscore <- function(y,x) { 
  tmp <- t.test(y~x)
  c(score=tmp$statistic,pval=tmp$p.value)
}
```

The next function is a wrapper to define a data matrix, X, and outcome
vector, Y, for different conditions. By default we can generate a
complete set of data from the null hypthesis to check the FWER. We can
also generate data with a subset of features which have different
means between groups to check, false positive rate, false negative
rate, and power by changing `propDiff` and `Diff`.

### Wrapper of data generation and differential analysis

```{r, wrapperDF}
wrapDA <- function(Nsamps, # Number of samples per pseudogroup
                   Nrow = 200, # Number of parameters
                   mean = 0, # Mean of normal
                   sd = 0.5, # SD of normal
                   alpha = 0.05, # Signficance threshold
                   propDiff = 0, # Proportion to make Ha
                   Diff = 0, # Difference for Ha features
                   reportAll = FALSE
                   )
  {
  
  # Genreate a data of random data from null distribution
  DAT <- matrix( rnorm( Nrow * Nsamps * 2, mean=0, sd=0.5), nrow=Nrow, ncol=( Nsamps *2) )
  
  # Add difference to one group
  if(propDiff > 0){
    nChange <- floor(propDiff*Nrow)
    DAT[1:(nChange), 1:Nsamps] <- DAT[1:(nChange), 1:Nsamps] + Diff
    rownames(DAT) <- paste0(c(rep("Ha", nChange), rep("H0", (Nrow - nChange))), 1:Nrow)
  }
  
  ## generate a (head/tail) phenotype of proper size
  pheno <- factor(rep(c('head','tail'),each=Nsamps))

  ## perform t.test on each data row with respect to the random phenotype
  DIF <- as.data.frame(t(apply(DAT,1,tscore,x=pheno)))
  
  ## Sort by p-value
  DIF <- DIF[order(DIF$pval),]
  
  ## Perform Bonferroni and BH corrections
  DIF$BFp <- p.adjust(DIF$pval, method = "bonferroni")
  DIF$BHp <- p.adjust(DIF$pval, method = "BH")
  
  ## Get BH Significant Rows
  DIFsig <- DIF[DIF$BHp < alpha,]
  
  ## Return All results or only features with small p-values
  if(reportAll){
    return(DIF)
  } else {
    ## If any genes BH significant return these rows else just return row with smallest p-value
    if(nrow(DIFsig)>0) return(DIFsig) else return(DIF[1,])
  }
}  
```

## Run differential analysis from null distributions for different sample sizes

First, we will check the FWER between Bonferroni and BH corrections for two different sample sizes: 10 samples per group and 50 samples per group. Each feature will have mean=0 and SD=0.5. For each run, the result will either be the feature with the smallest p-value, or a data.frame of features that passed BH significance threshold.

```{r}
sGroups <- c(10, 25, 50) # Vector of group sizes

# Run simulation for each sample size 100 times
outList <- lapply(sGroups, function(Nsamps){
    replicate(100, wrapDA(Nsamps), simplify = FALSE)
})

# Name list by sample size
names(outList) <- paste("Samps", sGroups, sep = "_")

# What does the output look like?
outList$Samps_10[[1]]
```


### Get family wise error rates from Bonferroni (alpha = 0.05)

#### Function to calculate FWER from list of experiment results

```{r}
getFWER <- function(resList, # List of lowest p-values
                    alpha = 0.05, # P-value threshold
                    pCol = "BFp" # Column with MHT P-value of interest
                    )
{
  ## Which experiments had at least one positive hit?
  FWERvec <- unlist(lapply(resList, function(y) sum(y[,"BFp"] < alpha)>0))
  
  ## What proportion of experiments had at least one positive hit?
  FWERval <- mean(FWERvec)
  
  ## Return FWER for this set of experiments
  return(FWERval)
}
```

```{r}
## Get FWER for bonferonni corrected p-values
BonVec <- unlist(lapply(outList, getFWER, alpha = 0.05, pCol = "BFp")) # Note pCol
print(BonVec)

## Get FWER for Benjamini-Hochberg corrected p-values
BhVec <- unlist(lapply(outList, getFWER, alpha = 0.05, pCol = "BHp")) # Note pCol
print(BhVec)

```

Note, for repeated experiments with multiple hypotheses the FWER
returned from Bonferroni and Benjamini-Hochberg is the same

## Run differential analysis from alternative distriubtions for different sample sizes

Next, we will create data where 25% of the genes differ between groups
by 0.5. In the output, true positive will have row names starting with
"Ha".

```{r}
haList <- lapply(sGroups, function(Nsamps){
    replicate(100, wrapDA(Nsamps, propDiff = 0.25, Diff = 0.5), simplify = FALSE)
})

## Name list by sample size
names(haList) <- paste("Samps", sGroups, sep = "_")

## What does the output look like?
haList$Samps_10[[1]]
```

The rows with the prefix, *Ha*, come from the distribution for which
the alternative hypthesis is true.

### Get False Positive and False Negative Rates

#### Function to calculate the false positive rate and false negative rates from list of experiment results

The a false positive rate will be calculated for each experiment
individually

```{r}
getFPR_FNR <- function(resList, # List of lowest p-values
                       alpha = 0.05, # P-value threshold
                       pCol = "BFp", # Column with MHT P-value of interest
                       nMH = 200, # Number of hypothesis tested
                       nHa = 50 # Number of true positives
                       )
{  
  ## Extract instances of null hypotheses with small p-values from each experiment
  H0list <- lapply(resList, function(y) y[!grepl("Ha", rownames(y)),])
  
  # Extract vector of false postive rates from each experiment
  FPvec <- unlist(lapply(H0list, function(y) sum(y[,pCol]<alpha)/(nMH-nHa)))
  
  # Extract instances of alternative hypotheses with small p-values from each experiment
  Halist <- lapply(resList, function(y) y[grepl("Ha", rownames(y)),])
  
  # Extract vector of false negative rates from each experiment
  FNvec <- 1 - unlist(lapply(Halist, function(y) sum(y[,pCol]<alpha)/(nHa)))
  
  # Create data.frame of results
  FP_FN <- data.frame(FalsePosi = FPvec, FalseNeg = FNvec)
  
  return(FP_FN)
}
```

### Bonferroni Corrected P-values

```{r}
## Get FP and FN
BonFP_FN <- lapply(haList, getFPR_FNR, alpha = 0.05, pCol = "BFp", nMH = 200, nHa = 50)

## Concatenate the results for each sample size
BonFP_FN <- do.call(rbind, BonFP_FN)
head(BonFP_FN)

## Add column for sample size identifier
BonFP_FN$SampleSize <- sub("[.][[:digit:]]*", "", rownames(BonFP_FN))
head(BonFP_FN)
```

### Benjamini Hochberg Corrected P-values

```{r}
## Get FP and FN
BhFP_FN <- lapply(haList, getFPR_FNR, alpha = 0.05, pCol = "BHp", nMH = 200, nHa = 50)

## Concatenate the results for each sample size
BhFP_FN <- do.call(rbind, BhFP_FN)
head(BhFP_FN)

## Add column for sample size identifier
BhFP_FN$SampleSize <- sub("[.][[:digit:]]*", "", rownames(BhFP_FN))
head(BhFP_FN)
```

### Plot Results

We will create boxplots of the results with ``ggplot2``.

```{r}
## Concatenate results between Bonferroni and Benjamini-Hochberg
FP_FN <- rbind(BonFP_FN, BhFP_FN)
FP_FN$correction <- rep(c("Bonferroni", "BH"), each = nrow(BonFP_FN)) # Add correction type
head(FP_FN)

## Create one column of False Positive and False Negative Rates
FP_FN <- gather(FP_FN, key = "statistic", value = "measure", 1:2)
head(FP_FN)

# Create plot of results
ggplot(FP_FN, aes(x = SampleSize, 
                        y = measure, 
                        fill = correction, 
                        colour = correction)) +
  geom_boxplot() +
  facet_wrap(~statistic, nrow = 2, scales = "free_y")

```

What do you notice about the nature of the performance of the
FWER-based Bonferroni correction, and FDR-based Benjamini-Hochberg
correction.
