---
title: "Quality Control"
author: "Stefano Monti"
output:
  html_document:
    theme: united
    toc: yes
    code_folding: show
    css: "../style/BS831.css" 
---

```{r include=FALSE}
knitr::opts_chunk$set(message=FALSE, warning=FALSE, eval=TRUE)
devtools::load_all(".")
library(Biobase)
```

```{r eval=FALSE}
library(BS831)
library(Biobase)
```

Here we show some basic QC plots to assess the quality of the data and
the nature of the distribution. For this purpose, we use a small
Breast Cancer dataset we have pre-uploaded to the `data/` folder.

## Import Breast Cancer Data

```{r table}
## breast cancer subtypes dataset from ArrayExpress
data(AEDAT.collapsed)

eset <- AEDAT.collapsed
emat <- exprs(eset)

## Assign rownames to emat that are gene annotations (These were removed by mistake)
rownames(emat) <- rownames(fData(eset))

## Table of disease states
print(table(eset$Characteristics.DiseaseState))

## simplify labels
eset$Characteristics.DiseaseState <- gsub(" breast cancer","",eset$Characteristics.DiseaseState)
print(table(eset$Characteristics.DiseaseState))
```

## Boxplots of gene expression values by sample

Here, we plot the distribution of genes within each sample as
boxplots. This is a quick check of how symmetric the distributions
are, and whether any 'biases' are apparent.

```{r boxplot, warning=FALSE, message=FALSE}
## Boxplot of expression values per sample (sorted and colored by disease state)
sampleCol <- c("coral","green","blue","purple")[as.factor(eset$Characteristics.DiseaseState)]
ord <- order( eset$Characteristics.DiseaseState )
boxplot(as.data.frame(exprs(eset)[,ord]),col=sampleCol[ord],pch="-")

## If you have a slow machine, you might want to subsample (random 1000 genes from the total)
## here, we are also taking the opportunity to better size (cex) and orient (las) the axis labels
set.seed(321) # for reproducible results
boxplot(as.data.frame(exprs(eset)[sample(nrow(emat),1000),ord]),col=sampleCol[ord],pch="-",
        las=2,cex.axis=.75)

## Fancier graphics with ggplot
library(reshape2)
library(ggplot2)

df <- data.frame(t(exprs(eset)))
df$sample <- sampleNames(eset)
df$diseasestate <- eset$Characteristics.DiseaseState
df.melt <- melt(df, id=c("sample","diseasestate"))

p1 <- ggplot(df.melt, aes(factor(sample), value))+
geom_boxplot(aes(fill = factor(diseasestate)), outlier.shape = 95, outlier.size = 1)+
xlab("sample")+ylab("expression")+
theme_bw()+
theme(axis.text.x  = element_blank()) + ggtitle("Boxplot of expression values by sample")
p1
```

## Testing the (log-)normality of the gene distributions

A lot of the statistical analyses performed on microarray data are
based on the assumption that the data is (log-)normally
distributed. Below, we show the use of a Kolmogorov-Smirnov (KS) test
to compare each gene's (log-)distribution to a standard normal.

```{r test.lognormal, warning=FALSE, message=FALSE}

## simple wrapper to have ks.test return the pair <statistic,p-value>
ks.pair <- function(X,Y) { tmp <- ks.test(X,Y); c(ks=tmp$statistic,p=tmp$p.value) }

## now let's apply it to each gene in the dataset
emat <- exprs(eset)
emat01 <- (emat - rowMeans(emat)) / apply(emat,1,sd)
KS <- t(apply(emat01,1,ks.pair,pnorm))

## let's plot the deviation from normality (given by -log10(p)) as a function of the gene's expression
plot(rowMeans(emat),-log10(KS[,"p"]),ylab="-log10(p)",xlab="mean(gene)")
abline(h=-log10(0.01),col="red")
idx <- -log10(KS[,"p"])>-log10(0.01)
points(rowMeans(emat)[idx],-log10(KS[idx,"p"]),col="red")

## let's compute a more conservative mht-corrected FDR, and plot that
KS <- cbind(KS,fdr=p.adjust(KS[,"p"]))
plot(rowMeans(emat),-log10(KS[,"fdr"]),ylab="-log10(FDR)",xlab="mean(gene)")
abline(h=-log10(0.01),col="red")
idx <- -log10(KS[,"fdr"])>-log10(0.01)
points(rowMeans(emat)[idx],-log10(KS[idx,"fdr"]),col="red")
```

Here, we show an alternative 'implementation' where, instead of
standardizing the data (emat), we pass to `pnorm` the gene mean and sd.

```{r test.lognormalAlt, warning=FALSE, message=FALSE}

## Rather than transforming the data to emat01, we can use the original data and add
## mean = mean(X), sd = sd(X) to the ks.test() function.
## This will feed the mean and sd of the gene-wise expression to the pnorm function
ks.pnorm <- function(X,Y) { tmp <- ks.test(X,pnorm, mean=mean(X), sd=sd(X)); c(ks=tmp$statistic,p=tmp$p.value) }

## now we can run as before, but with the emat object, not emat01
KS1 <- t(apply(emat,1,ks.pnorm))

## let's plot the deviation from normality (given by -log10(p)) as a function of the gene's expression
plot(rowMeans(emat),-log10(KS1[,"p"]),ylab="-log10(p)",xlab="mean(gene)")
abline(h=-log10(0.01),col="red")
idx <- -log10(KS1[,"p"])>-log10(0.01)
points(rowMeans(emat)[idx],-log10(KS1[idx,"p"]),col="red")

## let's compute a more conservative mht-corrected FDR, and plot that
KS1 <- cbind(KS1,fdr=p.adjust(KS[,"p"]))
plot(rowMeans(emat),-log10(KS1[,"fdr"]),ylab="-log10(FDR)",xlab="mean(gene)")
abline(h=-log10(0.01),col="red")
idx <- -log10(KS1[,"fdr"])>-log10(0.01)
points(rowMeans(emat)[idx],-log10(KS1[idx,"fdr"]),col="red")
 
# Compare the test statistics
plot(KS[,"ks.D"], KS1[,"ks.D"])
```

## Variation Filtering

We now show how to filter genes based on different measures of
variation: standrad deviation (SD) and its robust counterpart, the
median absolute deviation (MAD).

```{r variationfilter, warning=FALSE, message=FALSE}

## variation filter: rank genes according to that variation across samples
## method 1: filter by standard deviation
MEAN <- apply(emat, 1, mean)
STDEV <- apply(emat, 1, sd)
MAD <- apply(emat, 1, mad)
MEDIAN <- apply(emat, 1, median)

## highlight the top 4000 genes by stdev
plot(MEAN, STDEV, pch = ".", main = "mean vs. stdev (log2 data)")
top4Ksd <- order(STDEV, decreasing = T)[1:4000]
points(MEAN[top4Ksd], STDEV[top4Ksd], pch = ".", col = "red")
legend("topright", pch = 20, col = "red", legend = "top 4K genes by stdev")
eset.top4ksd <- eset[top4Ksd,]
#saveRDS(eset.top4ksd, file=file.path(OMPATH,"data/AEDAT.collapsed.sd4k.RDS"))

## unlogged data (multiplicative error), but plotted on logged axis
MEANUNLOGGED <- rowMeans(2^emat)
SDUNLOGGED <- apply(2^emat,1,sd)
plot(MEANUNLOGGED,SDUNLOGGED,pch=".",log="xy",main="mean vs. stdev (unlogged data)")
top4K <- order(SDUNLOGGED,decreasing=T)[1:4000]
points(MEANUNLOGGED[top4K],SDUNLOGGED[top4K],pch=".",col="red")
legend("topright",pch=20,col="red",legend="top 4K genes by stdev(unlogged)")

## method 2: filter by MAD (median absolute deviation)
MAD <- apply(emat, 1, mad)
MEDIAN <- apply(emat, 1, median)

## highlight top 4000 genes by MAD
plot(MEDIAN, MAD, pch = ".", main = "median vs. MAD (log2 data)")
top4Kmad <- order(MAD, decreasing = T)[1:4000]
points(MEDIAN[top4Kmad], MAD[top4Kmad], pch = ".", col = "red")
legend("topright", pch = 20, col = "red", legend = "top 4K genes by MAD")
eset.top4kmad <- eset[top4Kmad,]
#saveRDS(eset.top4kmad, file=file.path(OMPATH,"data/AEDAT.collapsed.mad4k.RDS"))

## let us look at the overlap between the two filters
length(intersect(top4Ksd,top4Kmad))
```
