---
title: "Data Acquisition and ExpressionSet Basics"
author: "Anthony Federico, Stefano Monti"
output:
  html_document:
    theme: united
    toc: yes
    code_folding: show
    css: "../style/BS831.css" 
---

In this module, we illustrate how to download gene expression datasets
from [GEO](http://www.ncbi.nlm.nih.gov/geo).

To this end, we will download and pre-process the gene expression data from the LOAD
(Late Onset Alzheimer Disease) study in
<a href="http://www.cell.com/fulltext/S0092-8674(13)00387-5">
[Zhang et al., Cell 2013]
</a>. The dataset consists of 690 samples corresponding to three brain
regions (cerebellum, prefrontal cortex, ) from 230 subjects, including 121 LOAD patients and 109 controls.

```{r include=FALSE}
knitr::opts_chunk$set(message=FALSE, warning=FALSE, eval=TRUE)
devtools::load_all(".")
require(GEOquery)
require(Biobase)
overwrite <- FALSE
```

We start by loading some required packages and functions.

```{r settings, eval=FALSE}
require(BS831)
require(GEOquery)
require(Biobase)
overwrite <- FALSE
```


# Download data from GEO

We start by downloading the data from GEO. We will then show two
approaches to adding gene annotation to the expression matrix, one
based on the use of already available annotation available on GEO, and
another one based on "de-novo" annotation based on the `biomaRt`
package.

```{r process.data1}
## Download expression data from GEO and store in temporary directory
tmp_dir <- tempdir()
LOAD <- getGEO(GEO="GSE44772", GSEMatrix=TRUE, destdir=tmp_dir)
LOAD <- LOAD[[1]] # getGEO returns a list, so we extract the first element
print(LOAD)       # display summary info for the object
```

## Use pre-existing Gene Annotation

We first show how to use already available gene annotation
information. To this end, we manually downloaded to the data/ subfolder the
[GPL4372.annot](https://www.ncbi.nlm.nih.gov/geo/query/acc.cgi?acc=GPL4372)
file from geo, then ran the following command.

```{r download.gene.annotation}
## read.delim can read compressed files
fdata <- read.delim(file=file.path(system.file("extdata", package="BS831"), "GPL4372.annot"), skip=27,row.names=1)
print(colnames(fdata))
```

There are many more annotation columns than needeed, thus we extract
some relevant columns, and since we are at it, we also properly sort
the rows, and rename the columns, for easier handling.

```{r process.data2}
fdata <- fdata[match.nona(featureNames(LOAD),rownames(fdata)),c("Gene.symbol","Gene.title","Gene.ID")]
## modifying fdata colnames for ease of handling
colnames(fdata) <- gsub("Gene.symbol","gene_symbol",colnames(fdata))
colnames(fdata) <- gsub("Gene.title","gene_title",colnames(fdata))
colnames(fdata) <- gsub("Gene.ID","GeneID",colnames(fdata)) 

## always be super-cautious, double- and triple-check
if ( any(featureNames(LOAD)!=rownames(fdata)) ) stop( "row mismatch" )

## finally, update the gene annotation in the ExpressionSet
LOAD1 <- LOAD
fData(LOAD1) <- fdata

## save data
if (overwrite) {
    saveRDS(LOAD1, file=file.path(OMPATH,"data/LOAD1.RDS"))
}
```

## Gene Annotation based on `biomaRt`

We will use the `EntrezGeneID` column in the fData to
retrieve gene symbols and descriptions with `biomaRt`.

```{r biomaRt}
## check the annotation columns in fData
print(colnames(fData(LOAD)))

## we first restrict to rows w/ non-empty EntrezGeneID annotation
LOAD <- LOAD[!is.na(fData(LOAD)[,"EntrezGeneID"]),]; nrow(LOAD)

## notice that there are replicate entries, but we will deal with them later
print( length(unique(fData(LOAD)[,"EntrezGeneID"])) )

## use biomaRt databased to retrieve the relevant annotation (type
## `?useMart` and `?getBM` for details)
mart <- biomaRt::useMart(biomart="ensembl", dataset="hsapiens_gene_ensembl")
martMap <- biomaRt::getBM(attributes=c("entrezgene_id","hgnc_symbol","description"),
                          filters="entrezgene_id",
                          values=fData(LOAD)[,"EntrezGeneID"],
                          mart=mart)

## remove entries with empty gene symbol
nrow(martMap) # before
martMap <- martMap[martMap[,"hgnc_symbol"]!="",]
nrow(martMap) # after

## notice that there are some duplicated entrez IDs and gene symbols
sum(base::duplicated(martMap[,"entrezgene_id"]))     
sum(base::duplicated(martMap[,"hgnc_symbol"]))

## we adopt the simple (perhaps too simple) approach of taking the
## first occurence of each EntrezID
nrow(martMap)
martMap <- martMap[match(unique(martMap[,"entrezgene_id"]),martMap[,"entrezgene_id"]),]
nrow(martMap)

## now there are no more replicated EntrezID (but there are still replicate gene symbols)
sum(base::duplicated(martMap[,"entrezgene_id"]))     
sum(base::duplicated(martMap[,"hgnc_symbol"]))

## let us now match gene annotation and expression data
matchIdx <- match.nona( martMap[,"entrezgene_id"], fData(LOAD)[,"EntrezGeneID"] )
LOAD2 <- LOAD[matchIdx,]
if ( any(martMap[,"entrezgene_id"]!=fData(LOAD2)[,"EntrezGeneID"]) ) stop( "row mismatch" )
fData(LOAD2) <- martMap

## save data
if (overwrite) {
    saveRDS( LOAD2, file=file.path(OMPATH,"data/LOAD2.RDS") )
}
```

## Merge Replicated Entries

This next step is not necessary, as one might want to keep the
multiple chip probes associated with the same gene symbol separate until
the end of the analysis. However, if desired, one can merge the rows
corresponding to multiple probes.

Here, we define a simple function to collapse multiple rows by
median. The function is based on the R functions `melt` and `dcast`
defined in the package `reshape2`.

```{r collapseByMedian.function}
require(reshape2)
collapseByMedian <- function(eset, rowid)
{
    ## require(reshape2)

    ## remove unmapped probe sets
    genes <- fData(eset)[, rowid]
    rows.mapped <- !is.na(genes) & genes != ""
    eset <- eset[rows.mapped,]
    genes <- fData(eset)[, rowid]

    ## collapse by median value among duplicate probes
    df <- data.frame(exprs(eset), genes = genes)
    df.melt <- melt(df, id.vars = "genes")
    df.median.collapsed <- dcast(df.melt, genes ~ variable, median)

    ## reassemble collapsed eset
    fdat <- fData(eset)
    if ( any(is.na(roworder <- match(df.median.collapsed[,'genes'],fdat[,rowid]))) )
        stop( "something wrong" )
    fdat.collapsed <- fdat[roworder,]
    eset <- ExpressionSet(assayData=as.matrix(df.median.collapsed[, colnames(eset)]),
                          phenoData=AnnotatedDataFrame(pData(eset)),
                          featureData=AnnotatedDataFrame(fdat.collapsed))
    return(eset)
}
collapseByFun <- function(eset,rowid,method=c("median","mean"))
{
    tbl <- data.frame(key=fData(eset)[,rowid],
                      exprs(eset))
    tbl1 <- tbl %>% group_by(key) %>% summarize_all(median)

    dat <- as.matrix(tbl1[,-match("key",colnames(tbl1))])
    rownames(dat) <- tbl1[,match("key",colnames(tbl1))]

    ExpressionSet(assayData=dat,
                  phenoData=pData(eset)[,match(colnames(dat),sampleNames(eset))],
                  featureData=fData(eset)[match(rownames(dat),fData(eset)[,rowid]),])
    
}
```

We now apply `collapseByMedian` to both
LOAD1 (n=`r sum(base::duplicated(fData(LOAD1)[,"gene_symbol"]))` duplicates), and 
LOAD2 (n=`r sum(base::duplicated(fData(LOAD2)[,"hgnc_symbol"]))` duplicates).
Notice that running it on a 600+ sample size is somewhat slow.

```{r collapse.load1, eval=FALSE}
nrow(LOAD1)
LOAD1.collapsed <- collapseByMedian(LOAD1, rowid="gene_symbol")
nrow(LOAD1.collapsed)
if (overwrite) {
    saveRDS(LOAD1.collapsed,file=file.path(OMPATH,"data/LOAD1.collapsed.RDS"))
}
tbl <- data.frame(key=c("a","b","a","b"),col1=1:4,col2=5:8)

```

```{r collapse.load2, eval=FALSE}
nrow(LOAD2)
LOAD2.collapsed <- collapseByMedian(LOAD2, rowid="hgnc_symbol")
nrow(LOAD2.collapsed)
if (overwrite) {
    saveRDS(LOAD2.collapsed,file=file.path(OMPATH,"data/LOAD2.collapsed.RDS"))
}
```
