---
title: "Mixture of Normals"
author: "Stefano Monti"
output:
  html_document:
    theme: united
    toc: yes
    code_folding: show
    css: "../style/BS831.css" 
---

```{r include=FALSE}
knitr::opts_chunk$set(message=FALSE, warning=FALSE, eval=TRUE)
devtools::load_all(".")
library(mclust)
```

```{r eval=FALSE}
library(BS831)
library(mclust)
```

Here we describe the definition and use of few simple functions to
generate data and to plot from a univariate mixture of Normal
distributions. According to R conventions, we define the four functions
`rmixnorm`, `pmixnorm`, `dmixnorm`, and `qmixnorm`. These functions
are defined in the file `mixnorm.R`. We also briefly introduce the
package [`Mclust`](https://www.stat.washington.edu/mclust/) (see [Quick
Tour](https://cran.r-project.org/web/packages/mclust/vignettes/mclust.html)).

## A simple mixnorm function

Below we define simple functions for sampling from, and computing the
density of, Gaussian mixtures. In particular, inspect the function
`rmixnorm` for the operational definition of the _Data Generating
Process_ discussed in class. We have also defined a function
`plot.mixnorm` (in `code/mixnorm.R`, not shown here) for plotting a
mixnorm's density function. See `Distributions.Rmd` for more details on how
to write a density plotting function.

```{r mixnorm.functions}
## show the code
print(rmixnorm) 
print(dmixnorm)
```

Below, we generate a sample from a 3-component Gaussian mixture, and
show the data histogram together with the density plot.

```{r rmixnorm}
set.seed(345) # for reproducible results
M3 <- rmixnorm(500,p=rep(1/3,3),mu=c(0,5,10),sigma=c(1,2,3))
hist(M3,probability=TRUE)
plot.mixnorm(1000,p=rep(1/3,3),mu=c(0,5,10),sigma=c(1,2,3),add=TRUE)

## Plot's truncated. Let's try the other way around ..
plot.mixnorm(1000,p=rep(1/3,3),mu=c(0,5,10),sigma=c(1,2,3))
hist(M3,probability=TRUE,add=TRUE) 
abline(v=c(0,5,10),lty=3,col='red')
```

## The MClust package

Here we briefly show the use of <a
href="http://www.stat.washington.edu/mclust/">MClust</a> (see <a
href="https://cran.r-project.org/web/packages/mclust/vignettes/mclust.html">vignette</a>),
a package that implements the fitting of Gaussian mixture models
(univariate and multivariate), and which uses Expectation Maximization
(EM) for parameter estimation, and the BIC approximation of the
marginal likelihood P(D|M) for model selection. We start by fitting
and testing models with between 1 and 6 mixture components with
unequal (i.e., cluster-specific) variances.

```{r mclust1}
MC3fit <- Mclust(M3,G=1:6,model="V")
summary(MC3fit)
print(MC3fit$parameters)

## boxplot the data as a function of their cluster assignment
boxplot(M3~MC3fit$classification)
```

Even with 500 points, because of the large variance of the 2nd and 3rd
components, the fitting procedure was not capable of guessing the
'correct' number of clusters. Let's try by either increasing the
sample size or decreasing the variance.

```{r mclust.two}
set.seed(123)
## increase the sample size
M3 <- rmixnorm(5000,p=rep(1/3,3),mu=c(0,5,10),sigma=c(1,2,3))
plot.mixnorm(5000,p=rep(1/3,3),mu=c(0,5,10),sigma=c(1,2,3),main="5,000 data points")
hist(M3,probability=TRUE,add=TRUE)
abline(v=c(0,5,10),lty=3,col='red')

MC3fit <- Mclust(M3,G=1:6,model="V")
summary(MC3fit)
print(MC3fit$parameters)

## boxplot the data as a function of their cluster assignment
boxplot(M3~MC3fit$classification)
abline(h=MC3fit$parameters$mean,col="red",lty=3)

## decrease the variance
set.seed(123)
M3 <- rmixnorm(500,p=rep(1/3,3),mu=c(0,5,10),sigma=c(1,1,1))
plot.mixnorm(500,p=rep(1/3,3),mu=c(0,5,10),sigma=c(1,1,1),main="500 data points")
hist(M3,probability=TRUE,add=TRUE)
abline(v=c(0,5,10),lty=3,col='red')

MC3fit <- Mclust(M3,G=1:6,model="V")
summary(MC3fit)
print(MC3fit$parameters)
```
