---
title: "Sample Size and Statistical Significance"
author: "Stefano Monti"
output:
  html_document:
    theme: united
    toc: yes
    code_folding: show
    css: "../style/BS831.css" 
---

In this module, we show how testing for multiple hypotheses (genes) can
increase the chance of false positives, especially for small sample
sizes. 

```{r include=FALSE}
knitr::opts_chunk$set(message=FALSE, warning=FALSE, eval=TRUE)
library(heatmap.plus)
```

```{r eval=FALSE}
library(heatmap.plus)
```

# Sample Size's effect on heatmap visualization

Here we show the scenario presented in class (slide "Gene markers
selection: better than chance?"). In the examples below, we show
heatmaps corresponding to random noise, and we show that, if enough
hypotheses are tested (in this case, 10,000), and the sample size is
sufficiently small (e.g., n=6), we can easily identify 'genes' whose
expression pattern _seems_ to be strongly associated with the
phenotype (in this case, a random head/tail), as suggested by the
heatmap with a clear blue-to-red pattern. As the sample size increases
(e.g., n=100), it is more difficult to be 'fooled', as the
corresponding heatmap shows a less clear blue-to-red pattern.

We start by generating a large [10,000x100] matrix filled with random
values drawn from a Gaussian distribution with mean=0 and stdev=0.5.

```{r random_matrix}
set.seed(123) # for reproducible results
Nrow <- 10000
Ncol <- 100
DAT <- matrix(rnorm(Ncol*Nrow,mean=0,sd=0.5),nrow=Nrow,ncol=Ncol)
hist(DAT)
```

We then pick a small subset of columns from this matrix and randomly
assign them a binary (0-1) phenotype. We then pick the top 25 markers
associated to 0 and 1, and plot the corresponding heatmaps.

```{r sample.size.heatmaps}
## creating a black-to-red palette for heatmap display
ramp.br <- colorRamp(c( "blue","white","red"))
palette.blue2red <- rgb( ramp.br(seq(0, 1, length = 14)), max = 255)

## wrapper function extracting t-statistic and p-value from a call to function t.test
tscore <- function(y,x) { 
  tmp <- t.test(y~x)
  c(score=tmp$statistic,pval=tmp$p.value)
}
## carry out multiple differential analysis on progressively larger sample sizes
for (ndraw in c(6,14,30,100) ) 
{ 
  ## randomly select Ncol columns from the full matrix
  DATi <- DAT[,colDraw <- sample(Ncol,size=ndraw)]
  ## generate a (head/tail) phenotype of proper size
  pheno <- factor(rep(c('head','tail'),each=ndraw/2))
  ## perform t.test on each data row with respect to the random phenotype
  DIFi <- t(apply(DATi,1,tscore,x=pheno))
  
  ## pick top 25 markers in each direction
  topMarkers <- c(order(DIFi[,1],decreasing=FALSE)[1:25],order(DIFi[,1],decreasing=TRUE)[1:25])
  ## visualize the corresponding heatmap of 50 markers
  CSC <- c('green','orange')[pheno] # green=head, orange=tail
  heatmap.plus(DATi[topMarkers,],col=palette.blue2red,Colv=NA,Rowv=NA,ColSideColors=cbind(CSC,CSC),
               main=paste("sample size =",ndraw))
  
  ## show the top markers (by p-value)
  print(head(cbind(DIFi,FDR=p.adjust(DIFi[,2],method="BH"))[topMarkers,]))
}  
```
